### 2017-02-06 Manuscript/SW_early_life_telomeres.Rmd
Eds

There was positive correlation between RTL measured from different samples taken at different time points during an individual’s life (Fig. 2A), but this was very weak (marginal R^2^ = `r round2(r.squaredGLMM(RTL_RTL1)['R2m'])`), and not significant (estimate = `r round3(fixef(RTL_RTL1)['RTL'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_RTL1)['RTL',],lessthan = F),collapse = ', ')`). Although both cross-sectional and longitudinal data indicated a general trend of telomere shortening with age, we found that RTL - measured across two samples taken from the same individuals over time - increased with age in `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our `r nrow(dd3)` $\Delta$RTL~*individual*~ measurements (Fig. 2A). To test whether increases in telomere length in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of the same samples to the variance observed among different samples of the same individual. We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig. 2B). Splitting the longitudinal data into instances of decreasing (i.e. $\Delta$RTL < 0) and increasing (i.e. $\Delta$RTL > 0) telomere length revealed that not only did we observe significantly greater decrease in RTL within individuals compared to within samples (Wilcoxon test: P `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also a significantly greater increase (P `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`; Fig. 2B).

>>>>

There was positive correlation between RTL measured from different samples taken at different time points during an individual’s life (Fig. 2A), but this was very weak (marginal R^2^ = `r round2(r.squaredGLMM(RTL_RTL1)['R2m'])`), and not significant (estimate = `r round3(fixef(RTL_RTL1)['RTL'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_RTL1)['RTL',],lessthan = F),collapse = ', ')`). Although both cross-sectional and longitudinal data indicated a general trend of telomere shortening with age, we found that RTL - measured across two samples taken from the same individuals over time - increased with age in `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our `r nrow(dd3)` $\Delta$RTL~*individual*~ measurements (Fig. 2A). To test whether increases in telomere length in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of the same samples to the variance observed among different samples of the same individual. We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig. 2B). Splitting the longitudinal data into instances of decreasing (i.e. $\Delta$RTL < 0) and increasing (i.e. $\Delta$RTL > 0) telomere length revealed that not only did we observe significantly greater decrease in RTL within individuals compared to within samples (Wilcoxon test: P `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also a significantly greater increase (P `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`; Fig. 2B). To better visualise how longitudinal telomere dynamics, including the likelihood of lengthening, varied with age, we tested how $\Delta$RTL changed with (log) age. We found that $\Delta$RTL increased with log age (estimate = `r round3(fixef(DeltaRTL_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(DeltaRTL_LogAge)['LogAge',],lessthan = F),collapse = ', ')`), again showing that
